Universal Characterization of Physical Variability by \\Shannon's $-\sum_ip_i\,\textrm{ln}\,p_i$ Formula

### 2019-11-19 Articles/2019-ShannonEntropy/Submissions/AJP/AJP-ShannonEntropyIsVariability.tex
Update AJP-ShannonEntropyIsVariability.tex

\begin{abstract}
	The Shannon entropy, one of the cornerstones of information theory, is widely used in physics, particularly in statistical mechanics. Yet its characterization and connection to physics remains vague, leaving ample room for misconceptions and misunderstanding. In this paper we will show that the Shannon entropy can be fully understood as measuring the variability of the elements within a distribution: it characterizes how much variation can be found within a collection of objects. We will see that it is the only indicator that is continuous and linear, that quantifies the number of yes/no questions (i.e.~bits) that are needed to identify an element within the distribution, and we will see how applying this concept to statistical mechanics in different ways lead to the Boltzman, Gibbs and von Neumann entropies.
\end{abstract}

@@

The Shannon entropy $-\sum p_i \log p_i$ is well established in such disparate fields as computer science[cite], physics[cite] and ecology[cite]. It is most often introduced in physics through statistical mechanics, and associated to vague notions, such as uncertainty[cite], knowledge[cite], lack of knowledge[cite] and disorder[cite]. These prevent a proper understanding of the concept and often lead to confusion and misunderstandings.

This paper aims to give a crisp characterization of the Shannon entropy that is intuitive and precise. The general idea is that it measures the variability of the elements within a distribution, which is a general concept applicable to many branches of science and it is independent of the notion of entropy of thermodynamics and statistical mechanics. Therefore we will use the term \emph{``Shannon variability''}, leaving ``entropy'' to the physical concept. We will show that the expression $- \sum_i p_i \log p_i$ is the only linear indicator of variability. It measures the number of questions one must ask to identify an element of the distribution, linking its use to information theory. We will study how the expression works over continuous variables, and show how phase space is special as it leaves the Shannon variability invariant under change of coordinates. We will turn to statistical mechanics and apply Shannon variability in different ways to recover connections to the Boltzmann, Gibbs and von Neumann entropies.

While some details presented here may be well known in a particular community, we find that they may be new to another. Therefore we assume little knowledge of the subject from the reader and include all details that we feel are important to properly understand the subject without confusion, including two standard derivations for the formula and examples that are common in computer science but rarely used in physics. We will briefly touch over a few of the common misconceptions.

@@

The general setting is the following. We have a set of elements $E = \{e_\alpha\}_{\alpha=1}^N$. In accordance with the comments made in our introduction, they can for example represent the galaxies in the Laniakea Supercluster, the animals in the Gal\'{a}pagos islands, the molecules in a box of gas given a particular macrostate, the different outcomes of spin measurement for a given quantum state, or the words in the complete works of Shakespeare. The choice of the particular set is driven by the interests and practices of scientists and their fields of study and therefore it is, in this sense, discretionary. Yet, it must be objective in at least one respect: once a choice is made the actual elements are the same for everybody. Once we arbitrarily decided we will study the galaxies in the Laniakea Supercluster, it is a matter of fact that the Milky Way will be included. This also means that, whatever properties those elements will have, they will represent a matter of fact about that set. In turn, this will make some choices arguably more appropriate or insightful than others. For example, in taxonomy the set of all animals that have feathers may be more appropriate than the set of all animals that can fly: the first includes only birds while the second includes most birds, a few mammals and a lot of insects.

@@

Once we have constructed our distribution, we will want to construct an indicator $H$ that quantifies how much variability the elements exhibit within the distribution. That is, we want to quantify the degree of diversity that the values can have within the distribution. To that end, we want to define some suitable requirements for $H$.

We may be tempted to use standard statistical quantities, like the range or the variance, but this is not possible. First, if the values associated with the bins are non-numerical (e.g. types of galaxies, words), such statistical quantities are not well defined. Second, relabeling the values (e.g. switching names, switching units, non-linearly changing coordinates) does not have an impact on variability of the elements, while statistical quantities will in general be affected. This tells us that our indicator cannot depend on $Q$ itself, but only on the weights $p_i$. That is, we require $H=H(p_i)$.\footnote{Note that for continuous quantities the weights are densities and are affected by the choice of $Q$: the unit is required to specify the numeric value (e.g.~$1 \% / $mm) and this will change under unit transformation (e.g.~$1000 \% /$m). Therefore, as the weights $p_i$ themselves depend on the unit, the indicator $H(p_i)$ will in general depend on the choice of $Q$. Though this is a source of additional confusion, for the purpose of defining a measure of variability it does not change things conceptually. We will discuss the problem in section \ref{sec_cont}.}

@@

As noted before, the level at which we describe the elements is not absolute and can change. For example, we may choose to group the animals in the Gal\'{a}pagos islands first by class (e.g. mammals, birds, reptiles) and then later refine the mammals by species. In this case, we would like $H$ to combine linearly with respect to the weights. That is, we want the variability of the overall distribution $H_T = H_C + p_M H_M$ to be the variability over the classes $H_C$ plus the variability of the mammals $H_M$ weighted by the fraction of mammals $p_M$. This also makes the quantity additive when combining two independent distributions.

To sum up, we have the following three requirements:
\begin{enumerate}
\item $H$ depends only on $p_i$ and it does so continuously
\item If $p_i=1/I$ then $H$ is a monotonically increasing function of $I$
\item Let $p_i$ and $q_j$ be the weights for two distributions respectively over $I$ and $J$ bins. Let $r_k$ be the distribution over $K=I+J-1$ bins constructed by expanding the $a^{th}$ bin of the first distribution using the second distribution. More specifically, let $1 \leq a \leq I$, then $r_k = \{p_1, p_2, ..., p_{a-1}, p_{a}q_1, p_{a}q_2, ..., p_{a}q_J, p_{a+1}, ..., p_I \}$. Then $H(r_k) = H(p_i) + p_{a} H(q_j)$.
\end{enumerate}
These are the same requirements Shannon put forth for his expression\cite{Shannon}, from which he showed to be the only possible choice: $H(p_i) = - \sum p_i \log(p_i)$. This expression quantifies the variability of the elements within the distribution, the variety of values one finds. The precise meaning of this variability is context dependent, as the choices of the elements and binning are not fixed and the meaning of the weights depends on what the distribution is describing. But this is true for any mathematical object: a real number may represent mass, color in the frequency spectrum, the total money supply, the half-life for an isotope, a probability and so on.

@@

The Shannon variability may represent knowledge in some cases, but not in general. Consider the following two cases:
\begin{enumerate}
\item There is 50\% chance you won one million dollars in the lottery and 50\% chance you won nothing.
\item There is 50\% chance you won one million dollars in the lottery and 50\% chance you won half a million dollars.
\end{enumerate}
The distribution in both cases is the same, two bins 50\% chance each, and so is the Shannon variability. Yet, you know more in the second case: you know you have won at least half a million dollars.

Unfortunately, entropy in general is often associated with vague characterizations like the two presented. It is said to represent uncertainty, knowledge, lack of knowledge or disorder depending on the authors, which leads to confusion and misunderstanding.  The characterization we have given of the Shannon variability measure, on the other hand, applies to all cases and leads naturally to the assumptions required to rederive it. Our characterization therefore is more fundamental. If we look at the galaxies in our universe, what variability do they exhibit in terms of their types? If we look at the animals in the Gal\'{a}pagos island, what variability is expressed in terms of their species? If we look at the molecules in a given macrostate, what variability do they express in terms of their microstates? If we look at the words in the complete works of Shakespeare, what variability do we find in his vocabulary?

@@

To understand what the numerical value represents, consider this example. Suppose we fix a distribution, say the animals in the Galap\'{a}gos islands binned by their respective species. Suppose I pick a specific animal from the set and you want to know its species. Suppose the only way for you to get that information is to ask a series of questions with only two possible answers, yes or no. How many questions would you have to ask? In other words, we are playing a game of Twenty Questions.

@@

If we have binary questions, the logarithms will be in base two and the unit for Shannon variability will be bits. It will indicate the average number of yes/no questions we need to identify an element within the dsitribution. In general, you can pick any base $b$: in base three we have ternary questions and trits, for ten we have questions with ten possible answers and digits. We can also pick a non-integer base, like the natural base for logarithms, and we will have nats. This is why the Shannon variability is fundamental in information theory, because it quantifies \emph{how much information is needed to transfer a value picked from a distribution}.

@@

For example, suppose the population of pets in a country is as follows:
\begin{table}[h]
\begin{tabular}{lr}
dogs & 27\% \\
cats & 48\% \\
fish & 10\% \\
bird & 8\% \\
small mammal & 4\% \\
reptile & 3\%
\end{tabular}
\end{table}

For the first question, we group cats on one side and everything else on the other, to form a 48/52 split. So we can ask, ``Is it a cat?". If the answer is yes, we finished. If not, we need to continue. We can group dogs on one side and everything else on the other to form a 27/25 split. So we can ask, ``Is it a dog?". If the answer is yes, we are done. If not, we continue. We can group fish with reptile and bird with small mammal to form a 13/14 split. So we can ask, ``Is it a fish or a reptile?". If the answer is yes, the followup question would be ``Is it a fish?". If the answer is no, the followup question would be ``Is it a bird?". With this scheme, the encoding, where 1 represents 'yes' and 0 represents 'no' to the each of the questions asked, is as follows:
\begin{itemize}
\item dogs 27\% - 2 questions (i.e.~2 bits) - answers: [no, yes] (i.e.~encoding 01)
\item cats 48\% - 1 question (i.e.~1 bit) - answers: [yes] (i.e.~encoding 1)
\item fish 10\% - 4 questions (i.e.~4 bits) - answers: [no, no, yes, yes] (i.e.~encoding 0011)
\item bird 8\% - 4 questions (i.e.~4 bits) - answers: [no, no, no, yes] (i.e.~encoding 0001)
\item small mammal 4\% - 4 questions (i.e.~4 bits) - answers: [no, no, no, no] (i.e.~encoding 0000)
\item reptile 3\% - 4 questions (i.e.~4 bits) - answers: [no, no, yes, no] (i.e.~encoding 0010)
\end{itemize}
The number of questions needed in each case corresponds to the number of bits. The answers in each case are represented by the encoding. Note how the encoding depends on the specific choice of questions. We can calculate the average number of bits for the encoding to be:  $(.48) * 1 + (.27) * 2 + (.1 + .08 + .04 + .03) * 4 = 2.02$ bits. This represents the average number of bits we would have to use for each animal if we repeated the game many times. We can also calculate the Shannon variability to be $.27 * \log_2(.27) + .48 * \log_2 (.48) + ... =1.98$ bits. This represents the ideal case, the minimum number of questions required to reach a definite answer. Note that our encoding is already very close to the ideal case.

@@

The use of information in physics, then, does not warrant a fundamental change of perspective in what constitutes a physical object, as some physicists have claimed. It is true that any physical process can be used to process information, \emph{when properly encoded}. It is also true that scientific theories, in the end, are models that can capture only the aspects of nature that can be tested experimentally, the information extracted by the experiment, under suitable circumstances. Therefore the claim that information plays an essential role in physical theories has a valid basis.\footnote{Comments like ``It is wrong, moreover, to regard this or that  physical quantity as sitting \emph{out there} with this or that numerical value it is absorbed'', ``the information thus solicited [by the experiment] make physics and comes in bits'' by Wheeler\cite{Wheeler} or ``I am proposing that the ultimate form of the implementable laws of physics requires only operations avaialble (in principle) in our actual universe'' by Landauer\cite{Landauer} go in this direction} But it is also true that that data requires the context to be able to be understood: we need to know what is the subject of our experiment, how to prepare it and how to collect the data. The art of experimental science is contained neither in the mathematical description nor in the data collected. As information, in the information theoretic sense, requires that context to become intelligible, it cannot play a primary role. Therefore the claim that the universe itself is information\footnote{Comments like ``[Information is] the fundamental building block of the Universe'' by Vedral or ``[Information] occupies the ontological basement'' by Davies go in this direction.} does not follow.

@@

Statistical mechanics aims to describe the collective behavior of a large number of physical systems (i.e.~particles), therefore asking what variability is expressed by such a collection is a well posed question. Yet, we have to understand that there are two distinct ways to apply the concept simply because there are two types of distributions in which one can be interested. We may consider the state of the whole system at a given time, and be interested in the distribution of the particles over all possible particle states. This will lead to the Boltzmann entropy. We may instead consider a statistical ensemble, which is a large collection of independent copies of the system found in different states, and consider the distribution of these different instances over the states of the whole system. This will lead to the Gibbs or the von Neumann entropy depending on whether the system is classical or quantum. Both these types of distributions are used and are therefore of interest, so we will look at both.

@@

Distributions over single-particle phase space is the setting used when deriving the Maxwell-Boltzmann distribution $\rho(q^i, p_i) = \left(\frac{\beta m}{2\pi} \right)^{3/2}e^{-\beta \frac{p_ip^i}{2m}}$, which is the distribution of particles for an ideal gas. As it evolves towards equilibrium, particles will spread out as much as they can under the constraints given by the energy, volume and number of particles, increasing the variability until is is maximized. The equilibrium is a statistical equilibrium, particles are moving around, but for any particle that moves in one direction, there is another one that moves in the opposite and the overall distribution remains the same.\footnote{This corresponds to the original insights developed by Boltzmann.}

@@

In this paper we have seen that:
\begin{itemize}
	\item the Shannon entropy measures the variability of the elements within a distribution, giving it a crisp intuitive meaning that is a general and applicable to all branches of science
	\item the expression is not arbitrary, as it is the only linear indicator for such concept
	\item it measures the variability by quantifying the number of questions one must ask to identify an element within the distribution, which corresponds to the number of bits needed to transmit or store that information
	\item when properly applied to statistical mechanics, the variability leads to the Boltzmann, Gibbs and von Neumann entropies.
\end{itemize}
The characterization we gave to the Shannon formula, then, is more precise than the common characterizations, such as disorder, information or lack of knowledge, and it should lead to less confusion. It clarify that the Shannon variability is an independence concept from the entropy of thermodynamics and statistical mechanics, and a link can be recovered only if properly applied. We find that this approach, once internalized, gives greater intuitive insight which also maps more readily with the mathematical details.

>>>>

\begin{abstract}
	The Shannon entropy, one of the cornerstones of information theory, is widely used in physics, particularly in statistical mechanics. Yet its characterization and connection to physics remain vague, leaving ample room for misconceptions and misunderstanding. We will show that the Shannon entropy can be fully understood as measuring the variability of the elements within a given distribution: it characterizes how much variation can be found within a collection of objects. We will see that it is the only indicator that is continuous and linear, that it quantifies the number of yes/no questions (i.e.~bits) that are needed to identify an element within the distribution, and we will see how applying this concept to statistical mechanics in different ways leads to the Boltzmann, Gibbs and von Neumann entropies.
\end{abstract}

@@

The Shannon entropy $-\sum p_i \log p_i$ is well established in such disparate fields as computer science[cite], physics[cite] and ecology[cite]. It is most often introduced in physics through statistical mechanics, and associated with vague notions, such as uncertainty[cite], knowledge[cite], lack of knowledge[cite] and disorder[cite]. These prevent a proper understanding of the concept and often lead to confusion.

This paper aims to give a crisp characterization of the Shannon entropy that is intuitive and precise. The general idea is that it measures the variability of the elements within a distribution, which is a general concept applicable to many branches of science, and it is independent of the notion of entropy of thermodynamics and statistical mechanics. Therefore we will use the term \emph{``Shannon variability,''} leaving ``entropy'' to the physical concept. We will show that the expression $- \sum_i p_i \log p_i$ is the only linear indicator of variability. It measures the number of questions one must ask to identify an element of the distribution, linking its use to information theory. We will study how the expression works over continuous variables, and show how phase space is special as it leaves the Shannon variability invariant under change of coordinates. We will turn to statistical mechanics and apply Shannon variability in different ways to recover connections to the Boltzmann, Gibbs and von Neumann entropies.

While some details presented here may be well known in a particular community, we find that they may be new to another. Therefore we assume little knowledge of the subject from the reader and include all details that we feel are important to properly understand the subject without confusion, including two standard derivations for the formula and examples that are common in computer science but rarely used in physics. We will briefly touch on a few of the common misconceptions.

@@

The general setting is the following. We have a set of elements $E = \{e_\alpha\}_{\alpha=1}^N$. In accordance with the comments made in the introduction, they can for example represent the galaxies in the Laniakea Supercluster, the animals in the Gal\'{a}pagos islands, the molecules in a box of gas given a particular macrostate, the different outcomes of spin measurement for a given quantum state, or the words in the complete works of Shakespeare. The choice of the particular set is driven by the interests and practices of scientists and their fields of study and therefore it is, in this sense, discretionary. Yet, it must be objective in at least one respect: once a choice is made the actual elements are the same for everybody. Once we arbitrarily decided we will study the galaxies in the Laniakea Supercluster, it is a matter of fact that the Milky Way will be included. This also means that, whatever properties those elements have, they will represent a matter of fact about that set. In turn, this will make some choices arguably more appropriate or insightful than others. For example, in taxonomy the set of all animals that have feathers may be more appropriate than the set of all animals that can fly: the first includes only birds while the second includes most birds, a few mammals and a lot of insects.

@@

Once we have constructed our distribution, we want to construct an indicator $H$ that quantifies how much variability the elements exhibit within the distribution. That is, we want to quantify the degree of diversity that the values can have within the distribution. To that end, we want to define some suitable requirements for $H$.

We may be tempted to use standard statistical quantities, like the range or the variance, but this is not possible. First, if the values associated with the bins are non-numerical (e.g. types of galaxies, words), such statistical quantities are not well defined. Second, relabeling the values (e.g.~switching names, switching units, non-linearly changing coordinates) does not have an impact on variability of the elements, while statistical quantities will in general be affected. This tells us that our indicator cannot depend on $Q$ itself, but only on the weights $p_i$. That is, we require $H=H(p_i)$.\footnote{Note that for continuous quantities the weights are densities and are affected by the choice of $Q$: the unit is required to specify the numeric value (e.g.~$1 \% / $mm) and this will change under unit transformation (e.g.~$1000 \% /$m). Therefore, as the weights $p_i$ themselves depend on the unit, the indicator $H(p_i)$ will in general depend on the choice of $Q$. Though this is a source of additional confusion, for the purpose of defining a measure of variability it does not change things conceptually. We discuss the issue in section \ref{sec_cont}.}

@@

As noted before, the level at which we describe the elements is not absolute and can change. For example, we may choose to group the animals in the Gal\'{a}pagos islands first by class (e.g.~mammals, birds, reptiles) and then later refine the mammals by species. In this case, we would like $H$ to combine linearly with respect to the weights. That is, we want the variability of the overall distribution $H_T = H_C + p_M H_M$ to be the variability over the classes $H_C$ plus the variability of the mammals $H_M$ weighted by the fraction of mammals $p_M$. This also makes the quantity additive when combining two independent distributions.

To sum up, we have the following three requirements:
\begin{enumerate}
\item $H$ depends only on $p_i$ and it does so continuously
\item If $p_i=1/I$ then $H$ is a monotonically increasing function of $I$
\item Let $p_i$ and $q_j$ be the weights for two distributions respectively over $I$ and $J$ bins. Let $r_k$ be the distribution over $K=I+J-1$ bins constructed by expanding the $a^{th}$ bin of the first distribution using the second distribution. More specifically, let $1 \leq a \leq I$, then $r_k = \{p_1, p_2, ..., p_{a-1}, p_{a}q_1, p_{a}q_2, ..., p_{a}q_J, p_{a+1}, ..., p_I \}$. Then $H(r_k) = H(p_i) + p_{a} H(q_j)$.
\end{enumerate}
These are the same requirements Shannon put forth for his expression\cite{Shannon}, from which he showed that the only possible choice is $H(p_i) = - \sum p_i \log(p_i)$. This expression quantifies the variability of the elements within the distribution, the variety of values one finds. The precise meaning of this variability is context dependent, as the choices of the elements and binning are not fixed and the meaning of the weights depends on what the distribution is describing. But this is true for any mathematical object: a real number may represent mass, color in the frequency spectrum, the total money supply, the half-life for an isotope, a probability and so on.

@@

The Shannon variability may represent knowledge in some cases, but not in general. Consider the following two cases:
\begin{enumerate}
\item There is 50\% chance you won one million dollars in the lottery and 50\% chance you won nothing.
\item There is 50\% chance you won one million dollars in the lottery and 50\% chance you won half a million dollars.
\end{enumerate}
The distribution in both cases is the same, two bins 50\% chance each, and so is the Shannon variability. Yet, you know more in the second case: you know you won at least half a million dollars.

Unfortunately, entropy in general is often associated with vague characterizations like the two presented. It is said to represent uncertainty, knowledge, lack of knowledge or disorder depending on the authors, which leads to confusion and misunderstanding.  The characterization we have given of the Shannon variability measure, on the other hand, applies to all cases and leads naturally to the assumptions required to rederive it. Our characterization therefore is more fundamental. If we look at the galaxies in our universe, what variability do they exhibit in terms of their types? If we look at the animals in the Gal\'{a}pagos islands, what variability is expressed in terms of their species? If we look at the molecules in a given macrostate, what variability do they express in terms of their microstates? If we look at the words in the complete works of Shakespeare, what variability do we find in his vocabulary?

@@

To understand what the numerical value represents, consider this example. Suppose we fix a distribution, say the animals in the Galap\'{a}gos islands binned by their respective species. Suppose we pick a specific animal from the set and you want to know its species. Suppose the only way for you to get that information is to ask a series of questions with only two possible answers, yes or no. How many questions would you have to ask? In other words, we are playing a game of Twenty Questions.

@@

If we have binary questions, the logarithms will be in base two and the unit for Shannon variability will be bits. It will indicate the average number of yes/no questions we need to identify an element within the dsitribution. In general, you can pick any base $b$: in base three we have ternary questions and trits, for ten we have questions with ten possible answers and digits. We can also pick a non-integer base, like the natural base for logarithms, and we will have nats. This is why the Shannon variability is fundamental in information theory, because it quantifies \emph{how much information is needed to transfer a value picked from a known distribution}.

@@

For example, suppose the population of pets in a country is as follows:
\begin{table}[h]
\begin{tabular}{lr}
dogs & 27\% \\
cats & 48\% \\
fish & 10\% \\
birds & 8\% \\
small mammals & 4\% \\
reptiles & 3\%
\end{tabular}
\end{table}

\noindent
For the first question, we group cats on one side and everything else on the other, to form a 48/52 split. So we can ask, ``Is it a cat?". If the answer is yes, we finished. If not, we need to continue. We can group dogs on one side and everything else on the other to form a 27/25 split. So we can ask, ``Is it a dog?". If the answer is yes, we are done. If not, we continue. We can group fish with reptiles and birds with small mammals to form a 13/14 split. So we can ask, ``Is it a fish or a reptile?". If the answer is yes, the followup question would be ``Is it a fish?". If the answer is no, the followup question would be ``Is it a bird?". With this scheme, the encoding, where 1 represents 'yes' and 0 represents 'no' to the each of the questions asked, is as follows:
\begin{itemize}
\item dogs 27\% - 2 questions (i.e.~2 bits) - answers: [no, yes] (i.e.~encoding 01)
\item cats 48\% - 1 question (i.e.~1 bit) - answers: [yes] (i.e.~encoding 1)
\item fish 10\% - 4 questions (i.e.~4 bits) - answers: [no, no, yes, yes] (i.e.~encoding 0011)
\item birds 8\% - 4 questions (i.e.~4 bits) - answers: [no, no, no, yes] (i.e.~encoding 0001)
\item small mammals 4\% - 4 questions (i.e.~4 bits) - answers: [no, no, no, no] (i.e.~encoding 0000)
\item reptiles 3\% - 4 questions (i.e.~4 bits) - answers: [no, no, yes, no] (i.e.~encoding 0010)
\end{itemize}
The number of questions needed in each case corresponds to the number of bits. The answers in each case are represented by the encoding. Note how the encoding depends on the specific choice of questions. We can calculate the average number of bits for the encoding to be:  $(.48) * 1 + (.27) * 2 + (.1 + .08 + .04 + .03) * 4 = 2.02$ bits. This represents the average number of bits we would have to use for each animal if we repeated the game many times. We can also calculate the Shannon variability to be $.27 * \log_2(.27) + .48 * \log_2 (.48) + ... =1.98$ bits. This represents the ideal case, the minimum number of questions required to reach a definite answer. Note that our encoding is already very close to the ideal case.

@@

The use of information in physics, then, does not warrant a fundamental change of perspective in what constitutes a physical object, as some physicists have claimed. It is true that any physical process can be used to process information, \emph{when properly encoded}. It is also true that scientific theories, in the end, are models that can capture only the aspects of nature that can be tested experimentally, the information extracted by the experiment, under suitable circumstances. Therefore the claim that information plays an essential role in physical theories has a valid basis.\footnote{Comments like ``It is wrong, moreover, to regard this or that  physical quantity as sitting \emph{out there} with this or that numerical value it is absorbed'', ``the information thus solicited [by the experiment] make physics and comes in bits'' by Wheeler\cite{Wheeler} or ``I am proposing that the ultimate form of the implementable laws of physics requires only operations avaialble (in principle) in our actual universe'' by Landauer\cite{Landauer} go in this direction} But it is also true that that data requires the context in order to be understood: we need to know what the subject of our experiment is, how to prepare it and how to collect the data. The art of experimental science is contained neither in the mathematical description nor in the data collected. As information, in the information theoretic sense, requires that context to become intelligible, it cannot play a primary role. Therefore the claim that the universe itself is information\footnote{Comments like ``[Information is] the fundamental building block of the Universe'' by Vedral or ``[Information] occupies the ontological basement'' by Davies go in this direction.} does not follow.

@@

Statistical mechanics aims to describe the collective behavior of a large number of physical systems (i.e.~particles), therefore asking what variability is expressed by such a collection is a well posed question. Yet, we have to understand that there are two distinct ways to apply the concept simply because there are two types of distributions in which one can be interested. We may consider the state of the whole system at a given time, and be interested in the distribution of the particles over all possible particle states. This will lead to the Boltzmann entropy. We may instead consider a statistical ensemble, which is a large collection of independent copies of the system found in different states, and consider the distribution of these different instances over the states of the whole system. This will lead to the Gibbs or the von Neumann entropy depending on whether the system is classical or quantum. Both these types of distributions are used and are therefore of interest, so we will examine both.

@@

Distributions over single-particle phase space constitute the setting used when deriving the Maxwell-Boltzmann distribution $\rho(q^i, p_i) = \left(\frac{\beta m}{2\pi} \right)^{3/2}e^{-\beta \frac{p_ip^i}{2m}}$, which is the distribution of particles for an ideal gas. As it evolves towards equilibrium, particles will spread out as much as they can under the constraints given by the energy, volume and number of particles, increasing the variability until it is maximized. The equilibrium is a statistical equilibrium, particles are moving around, but for any particle that moves in one direction, there is another one that moves in the opposite and the overall distribution remains the same.\footnote{This corresponds to the original insights developed by Boltzmann.}

@@

In this paper we have seen that:
\begin{itemize}
	\item the Shannon entropy measures the variability of the elements within a given distribution, giving it a crisp intuitive meaning that is a general and applicable to all branches of science
	\item the expression is not arbitrary, as it is the only linear indicator for such concept
	\item it measures the variability by quantifying the number of yes/no questions one must ask to identify an element within the distribution, which corresponds to the number of bits needed to transmit or store that information
	\item when properly applied to statistical mechanics, the variability leads to the Boltzmann, Gibbs and von Neumann entropies.
\end{itemize}
The characterization we gave to the Shannon formula, then, is more precise than the common characterizations, such as disorder, information or lack of knowledge, and it should lead to less confusion. It clarifies that the Shannon variability is an independent concept from the entropy of thermodynamics and statistical mechanics, and a link can be recovered only if properly applied. We find that this approach, once internalized, gives greater intuitive insight and also maps more readily to the mathematical details.
